Predicting the effects of character displacement on food-web dynamics

### 2019-03-07 manuscript_ECD_model.Rmd
add edits to Methods

Ecological character displacement is thought to be a key evolutionary process generating biodiversity [@Schluter2000; @Pfennig2010; but see @Stuart2013]. Ecological character displacement describes the “process of phenotypic evolution in a species generated or maintained by [exploitative] resource competition with one or more coexisting species” [@Schluter2000]. Over the past 40 years, a large body of theoretical [e.g. @Lawlor1976; @Abrams1986; @Doebeli1996; @Taper1985] and empirical [reviewed in: @Schluter2000; @Dayan2005; @Stuart2013] work has been generated to understand the scenarios under which exploitative competition for resources leads to the divergence of consumer foraging traits. The general result emerging from these studies is that, if resources are nutritionally substitutable [@Abrams1987; @Fox2008] and there is no other strong source of density dependence acting on consumers [@Abrams1986], then resource competition drives the divergence of consumer foraging traits [@Lawlor1976; @Taper1985]. This process is not simply driven by ecological differences, but creates an eco-evolutionary feedback that drives further differentiation. This key insight was made by theoretical models that explicitly included resource dynamics as a mediator of competition in driving evolutionary change [@Lawlor1976; @Abrams1986; @Taper1985].

Although explicit models of resources gave insights to the evolution of character displacement, the ecological feedback onto consumer-resource dynamics have received surprisingly little attention. 
This is likely because the ecological feedback has been primarily studied through the lens of competition and coexistence theory [@Lawlor1976; @Germain2018; @Bassar2017; @McPeek2019]. 
For example, early theoretical work by @Lawlor1976 showed that ecological character displacement promotes coexistence by favoring specialized consumers that experience reduced interspecific competition. 
Yet, this reduction in interspecific competition may, at the same time, increase the interspecific interaction between a consumer and the resource it specializes on.  
Both food-web theory and empirical studies have shown that increasing the strength of consumer-resource interactions often leads to resource suppression, which if sufficient enough, can generate oscillations, and hence less stable consumer-resource dynamics [@Rosenzweig1971; @Luckinbill1973; @Murdoch2002; @Murdoch2003; @McCann2011].
Thus, a food-web perspective, which accounts for both the direct and indirect effects of consumer-resource interactions, may yield new insight to the ecological consequences of character displacement.

Here, I address this knowledge gap by building a mathematical model to examine how ecological character displacement affects consumer-resource dynamics in a food-web context. I address two questions: (1) How does ecological character displacement affect resource abundances? (2) How does character displacement affect food-web stability? To test the generality of these ecological effects, I examined three different foraging scenarios: *CONDENSE THIS. NO NEED TO GIVE ALL THE DETAILS YET. RIGHT NOW IT IS JUST CONFUSING.* (i) resources occur in the same habitat and consumers exhibit a linear functional response [e.g. @MacArthur1972; @Lawlor1976]; (ii) resources occur in different habitats and consumers exhibit a linear functional response [e.g. @Lawlor1976]; (iii) resources occur in distinct habitats and consumers exhibit a more realistic functional response [@McCann2005]. I also explore the contingency of these effects to different assumptions about evolutionary trade-offs in consumer foraging traits. I focus on the ecological side of these dynamics because the evolutionary aspects have been well characterized [@Lawlor1976; @Taper1985; @Abrams1986; @McPeek2019]. I show that the consequences of character displacement on food-web dynamics are not always intuitive and depend on the foraging scenario and tradeoff in consumer traits.

# Methods and Results

## Underlying Food-web Dynamics

@@

where $r_i$ represents the intrinsic growth rate of resource $i$, $K_i$ represents the carrying capacity of resource $i$, $e_{ji}$ represents the conversion efficiency of resource $i$ into consumer $j$, and $m_j$ represents the mortality rate of consumer $j$. $F_{ji}(R_i)$ represents consumer $j$'s feeding rate (i.e. functional response) on resource $i$. This model is a useful characterization of a scenario where consumers compete for two distinct resources (e.g. zooplankton and benthic invertebrates) rather than if resources are better characterized by a continuous trait distribution (e.g. seed size, see @Taper1985 for an example). In the simplest scenario, consumers can forage for both resources simultaneously and their feeding rates increase linearly with resource density, such that $F_{ji}(R_i)=a_{ji}R_i$ where $a_{ji}$ is the attack rate of consumer $j$ on resource $i$ [@MacArthur1972]. The evolution of consumer attack rates in this model (and several extensions) have been analyzed in detail [@Lawlor1976; @Abrams1986], with the general result being divergent character displacement. I say consumers have undergone divergent character displacement if there evolved attack rates are more specialized, measured as $|\frac{a_{ii}}{a_{ii}+a_{ij}}-0.5|$, when evolving with a competing consumer. I focus here on the ecological consequences of this adaptive divergence.

To investigate the ecological consequences, I studied the effect of adaptive divergence in consumers on resource densities and the stability of the entire food web at equilibrium. An equilibrium is reached when the four rates of change in the above equations are 0, and solving the system at this point gives equilibrium densities for each resource ($\hat R_1,\hat R_2$) and consumer ($\hat C_1,\hat C_2$). At the coexistence equilibrium (i.e., where $\hat R_1,\hat R_2, \hat C_1,\hat C_2 > 0$), I calculated resource densities and performed a linear stability analysis [@Otto2007]. This stability analysis derives the dominant (largest in absolute value) eigenvalue, $\lambda$, which determines whether (and how readily) the food web, when perturbed a small amount, will return to equilibrium (see supplementary material for derivation). Together, these two measures indicate how ecological character displacement affects resource densities and food-web stability.

I examine the effects of character displacement on food-web dynamics by (i) deriving analytical expressions for the relationship between consumer attack rates and food-web dynamics and (ii) simulating the effects of competition on the eco-evolutionary dynamics of consumers and resources. To gain analytical insight, I assume that resources are equivalent ($r=r_1=r_2$ and $K=K_1=K_2$) as well as consumers ($e=e_{11}=e_{12}=e_{21}=e_{22}$; $m=m_1=m_2$), except that consumer attack rates are mirror images of each other ($a_{ii}=a_{11}=a_{22}$; $a_{ij}=a_{12}=a_{21}$). While this scenario is arguably simplistic, it allows me to isolate the effects of ecological character displacement.

To simulate eco-evolutionary dynamics, I used an Adaptive Dynamics approach. At each evolutionary time step, I created a mutant consumer by randomly choosing a consumer and modifying its attack rate on one resource by either subtracting or adding a small constant (0.001 in the following simulations) with equal probability. The mutant's attack rate on the other resource was determined by a tradeoff, such that $\big(a_{ii}/{A}\big)^n+\big(a_{ij}/{A}\big)^n=1$, where $A$ is the total investment in attack rates and $n$ describes the shape of the tradeoff [@Sargent2006]. This function has the useful property that it differentiates between cases where intermediate combinations of $a_{ii}$ and $a_{ij}$ are higher, on average, than the extremes (when $n>1$, green line in Fig. \ref{fig:MacArthur_Plot}A) or, conversely, where the two extremes are higher, on average, than intermediate investments (when $n<1$, orange line Fig. \ref{fig:MacArthur_Plot}A). When $n=1$, the tradeoff function is linear, and all combinations of $a_{ii}$ and $a_{ij}$ have the same total attack rate (blue line in Fig. \ref{fig:MacArthur_Plot}A). Assuming the mutant consumer was rare, I then determined whether the mutant had higher relative fitness than the resident consumer, and thus could invade and replace the resident consumer. If the mutant was able to invade, I updated the attack rate of the resident consumer to the mutant attack rate and allowed consumer and resource dynamics to reach a steady state. I then repeated the simulation up to 10,000 times, which was sufficient for consumers to either reach an evolutionary stable strategy [ESS, @Smith1973] or an evolutionary limit (e.g. $\frac{a_{ii}}{a_{ii}+a_{ij}}$ is constrained to a maximum of 1 and minimum of 0).

>>>>

Ecological character displacement is an important adaptive process in generating biodiversity [@Schluter2000; @Pfennig2010; but see @Stuart2013]. 
This process results from "phenotypic evolution in a species generated or maintained by [exploitative] resource competition with one or more coexisting species” [@Schluter2000]. 
Over the past 40 years, a large body of theoretical [e.g. @Lawlor1976; @Abrams1986; @Doebeli1996; @Taper1985; @McPeek2019] and empirical [reviewed in: @Schluter2000; @Dayan2005; @Stuart2013] work has examined which scenarios lead to phenotypic divergence or convergence of competing consumers.
The general result has been that, if resources are nutritionally substitutable [@Abrams1987; @Fox2008] and there is no other strong source of density dependence acting on consumers [@Abrams1986], then resource competition drives the adaptive divergence of competitors [@Lawlor1976; @Taper1985]. 
This adaptive process is not simply a response to statitc differences in resource distributions, but creates an eco-evolutionary feedback that drives further differentiation. 
This crucial insight was made by theoretical models that explicitly included resource dynamics as a mediator of competition in driving evolutionary change [@Lawlor1976; @Abrams1986; @Taper1985].

Although models that included resources gave insights to the evolution of character displacement, the ecological feedback onto consumer-resource dynamics has received surprisingly little attention. 
This is likely because the ecological feedback has been primarily studied through the lens of competition theory [@Lawlor1976; @Germain2018; @Bassar2017; @McPeek2019]. 
For example, early theoretical work by @Lawlor1976 showed that ecological character displacement promotes coexistence by favoring specialized consumers that experience reduced interspecific competition. 
Yet, this reduction in interspecific competition may, at the same time, increase interspecific interactions between specialized consumers and their resources.  
Both food-web theory and empirical studies have shown that increasing the strength of consumer-resource interactions often suppresses the abundance of resources, which if sufficient enough, can generate oscillations and less stable consumer-resource dynamics [@Rosenzweig1971; @Luckinbill1973; @Murdoch2002; @Murdoch2003; @McCann2011].
Thus, a food-web perspective, which accounts for both the direct and indirect effects of consumer-resource interactions, may yield new insight to the ecological consequences of character displacement.

Here, I address this knowledge gap by studying a mathematical model that examines how ecological character displacement affects consumer-resource dynamics in a food-web context. I address two questions: (1) How does ecological character displacement affect resource abundances? (2) How does character displacement affect food-web stability? To test the generality of these effects, I explored different assumptions of ecological foraging scenarios and evolutionary trade-offs in consumer foraging traits.

# Methods

@@

where $r_i$ represents the intrinsic growth rate of resource $i$, $K_i$ represents the carrying capacity of resource $i$, $e_{ji}$ represents the conversion efficiency of resource $i$ into consumer $j$, and $m_j$ represents the mortality rate of consumer $j$. $F_{ji}(R_i)$ represents consumer $j$'s feeding rate (i.e. functional response) on resource $i$. 
This model is a useful characterization of a scenario where consumers compete for two distinct resources (e.g. zooplankton and benthic invertebrates) rather than if resources are better characterized by a continuous trait distribution (e.g. seed size, see @Taper1985 for an example). 
In the simplest scenario, consumers can forage for both resources simultaneously and their feeding rates increase linearly with resource density, such that $F_{ji}(R_i)=a_{ji}R_i$ where $a_{ji}$ is the attack rate of consumer $j$ on resource $i$ [@MacArthur1972]. 
The evolution of consumer attack rates in this model (and several extensions) have been analyzed in detail [@Lawlor1976; @Abrams1986], with the general result being divergent character displacement. 
I say consumers have undergone divergent character displacement if there evolved attack rates are more specialized, measured as $|\frac{a_{ii}}{a_{ii}+a_{ij}}-0.5|$, when evolving with a competing consumer.

To investigate the ecological consequences of divergent character displacement, I analyzed differences in resource abundances and food-web stability at equilibrium. 
An equilibrium is reached when the four rates of change in the above equations are 0, and solving the system at this point gives equilibrium densities for each resource ($\hat R_1,\hat R_2$) and consumer ($\hat C_1,\hat C_2$). 
At the coexistence equilibrium (i.e., where $\hat R_1,\hat R_2, \hat C_1,\hat C_2 > 0$), I calculated resource densities and performed a linear stability analysis [@Otto2007]. 
This stability analysis derives the dominant eigenvalue, $\lambda$, which determines whether (and how readily) the food web will return to equilibrium after a small perturbation.

When possible, I derived analytical expressions for the relationship between consumer attack rates and food-web dynamics. 
To do this, I simplified the model by assuming that resources are equivalent ($r=r_1=r_2$ and $K=K_1=K_2$) as well as consumers ($e=e_{11}=e_{12}=e_{21}=e_{22}$; $m=m_1=m_2$), except that consumer attack rates are mirror images of each other ($a_{ii}=a_{11}=a_{22}$; $a_{ij}=a_{12}=a_{21}$). 
Controlling for other sources of variability allowed me to isolate the general effects of divergent character displacement on food-web dynamics.

Although divergent character displacement occurs in all scenarios we examined, the magnitude of this divergence may be context dependent.
Therefore, I simulated the eco-evolutionary feedback between consumers and resources using an Adaptive Dynamics approach.
At each evolutionary time step, I created a mutant consumer by randomly choosing a consumer and modifying its attack rate on one resource by either subtracting or adding a small constant (0.001 in the following simulations) with equal probability. 
The mutant's attack rate on the other resource was determined by a tradeoff, such that $\big(a_{ii}/{A}\big)^n+\big(a_{ij}/{A}\big)^n=1$, where $A$ is the total investment in attack rates and $n$ describes the shape of the tradeoff [@Sargent2006]. 
This function has the useful property that it differentiates between cases where intermediate combinations of $a_{ii}$ and $a_{ij}$ are higher, on average, than the extremes (when $n>1$, green line in Fig. \ref{fig:MacArthur_Plot}A) or, conversely, where the two extremes are higher, on average, than intermediate investments (when $n<1$, orange line Fig. \ref{fig:MacArthur_Plot}A). 
When $n=1$, the tradeoff function is linear, and all combinations of $a_{ii}$ and $a_{ij}$ have the same total attack rate (blue line in Fig. \ref{fig:MacArthur_Plot}A). 
Assuming the mutant consumer was rare, I then determined whether the mutant had higher relative fitness than the resident consumer, and thus could invade and replace the resident consumer. 
If the mutant was able to invade, I updated the attack rate of the resident consumer to the mutant attack rate and allowed consumer and resource dynamics to reach a steady state. 
I then repeated the simulation up to 10,000 times, which was sufficient for consumers to either reach an evolutionary stable strategy [ESS, @Smith1973] or an evolutionary limit (e.g. $\frac{a_{ii}}{a_{ii}+a_{ij}}$ is constrained to a maximum of 1 and minimum of 0).
